---
title: "Correlações Entre Miocinas e Dados Somatométricos"
author: "Leonardo R. Gama - leonardo.gama@usp.br"
date: "15/08/2020"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: false
    code_folding: hide
    df_print: paged
---

<!--
   - MIT License
   -
   - Copyright 2021 Alexandre Ferreira Ramos
   -
   - Permission is hereby granted, free of charge, to any person obtaining a
   - copy of this software and associated documentation files (the "Software"),
   - to deal in the Software without restriction, including without limitation
   - the rights to use, copy, modify, merge, publish, distribute, sublicense,
   - and/or sell copies of the Software, and to permit persons to whom the
   - Software is furnished to do so, subject to the following conditions:
   -
   - The above copyright notice and this permission notice shall be included in
   - all copies or substantial portions of the Software.
   -
   - THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR
   - IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY,
   - FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL
   - THE AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER
   - LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING
   - FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER
   - DEALINGS IN THE SOFTWARE.
   -->

<!--
   - Project:  ClinNutr2021 - AMPhyBio Laboratory <github.com/amphybio/ClinNutr2021>
   - Authors:  Leonardo Gama <leonardo.gama@usp.br>
   - Version:  1.0
   - Created:  15-08-2020
   - Updated:  05-06-2021
   - Compile:  Rscript -e 'rmarkdown::render("correlations.Rmd")'
   -->

```{r setup, include=FALSE}
library(knitr)
library(rprojroot)
opts_knit$set(root.dir=find_root(has_file('LICENSE.txt')))
opts_chunk$set(cache=TRUE, warning=FALSE, fig.align='center', fig.dim=c(10, 13))
options(width=100)
```

```{r imports, include=FALSE}
pkgs <- c(
          'autoimage',      # color scale legend
          'magrittr',       # pipe operator (%>%)
          'RColorBrewer'    # color palettes
         )

not_installed <- !pkgs %in% rownames(installed.packages())
if (any(not_installed)) install.packages(pkgs[not_installed])
lapply(pkgs, library, character=TRUE)

data(myokines)
```

```{r subset, include=FALSE}
# Select myokines columns.
i <- EXCEL_COLS[c('AA', 'AP')]
myokines <- names(raw_dat)[i[1]:i[2]]

# Reorder.
myok_ord <- c('FABP3', 'FSTL', 'BDNF', 'IL', 'Irisin', 'Myostatin')
myokines <- unlist(lapply(myok_ord, grep, myokines, value=TRUE))
site_ord <- c('muscle', 'plasma|serum', 'tumor')
myokines <- unlist(lapply(site_ord, grep, myokines, value=TRUE))

# Log transformation.
myok_dat <- raw_dat[, myokines] %>% lapply(log10) %>% data.frame

# Pretty labels.
myokines <- myokines %>% sub(' |_', ' (', .) %>% paste0(')')
names(myok_dat) <- myokines

# Select somatometric and inflamatory markers columns.
i <- EXCEL_COLS[c('E', 'H', 'AY', 'BC', 'BF', 'O', 'Q', 'R', 'X', 'Y', 'Z')]
somato <- names(raw_dat)[c(i[1]:i[2], i[3], i[4]:i[5], i[6:11])]
somato_dat <- raw_dat[, somato]

# Build data.frame with relevant variables.  
group <- raw_dat[[ grep('group', names(raw_dat))[1] ]]
bmi <- raw_dat[[ grep('BMI', names(raw_dat)) ]]
dat <- cbind(group, bmi, myok_dat, somato_dat)
cases <- (group == 'CC')
```

## Cálculos

Coeficientes parciais de correlação de Spearman ajustados para IMC em ambos os
grupos entre miocinas e dados antropométricos - peso atual, perda de peso,
composição corporal (músculo esquelético, tecido adiposo visceral, tecido
adiposo subcutâneo). Pedido do revisor:

>_Please provide in both patients:_
>
>  1. _a correlation matrix between myokines and somatometric data, as well as_
>  2. _a correlation matrix between myokines and somatometric data using partial
>  **Spearman's** correlation coefficients adjusted for BMI for both patients._

```{r calculation}
spearman_conf_int <- function(rho, n, conf.level=0.95) {
    # Calculate two-sided confidence interval for Spearman's rank correlation
    # using Fisher transformation, with a standard error for z as estimated by
    # Fieller et al. (1957). Code based on cor.test() for Pearson's correlation.
    sigma <- 1.03/sqrt(n - 3)
    z <- atanh(rho)
    ci <-  z + c(-1, 1) * sigma * qnorm((1 + conf.level)/2)
    ci <- tanh(ci)
    attr(ci, 'conf.level') <- conf.level
    return(ci)
}

par_cor <- function(x, y, covar, data, subset, method=c('pearson', 'spearman'), conf.level=0.95) {
    # Calculate partial correlation of 'x' and 'y' with 'covar' as covariate.
    # Statisticts are for the two-sided alternative hypothesis.
    method <- match.arg(method)
    if (!missing(data)) {
        x <- data[, x]
        y <- data[, y]
        covar <- data[, covar]
    }
    if (!missing(subset)) {
        x <- x[subset]
        y <- y[subset]
        covar <- covar[subset]
    }

    # Remove samples with missing data.
    ok <- complete.cases(x, y, covar)
    x <- x[ok]
    y <- y[ok]
    covar <- covar[ok]

    mean_x <- mean(x)
    mean_y <- mean(y)

    # Adjust by the covariate.
    x <- lm(x ~ covar)$residuals
    y <- lm(y ~ covar)$residuals

    # Correlation and confidence interval.
    res <- suppressWarnings(cor.test(x, y, method=method, conf.level=conf.level))
    res$n <- length(x)
    if (method == 'spearman')
        res$conf.int <- spearman_conf_int(res$estimate, res$n, conf.level)

    res$x <- x + mean_x
    res$y <- y + mean_y
    return(res)
}

groups <- c('cases', 'control')
methods <- c('pearson', 'spearman')
types <- c('unadjusted', 'adjusted')

# Calculate correlations for all possible scenarios.
cors <- list()
for (group in groups) for (method in 'spearman') for (type in types) {
    i <- paste(group, method, type, sep='_')
    cors[[i]] <- length(myokines) %>% replicate(list()) %>% setNames(myokines)
    subset <- if (group == 'cases') cases else !cases
    for (m in myokines) for (s in somato) {
        if (type == 'unadjusted') {
            x <- dat[subset, m]
            y <- dat[subset, s]
            res <- cor.test(x, y, method=method)
            res$x <- x
            res$y <- y
            res$n <- sum(complete.cases(x, y))
            if (method == 'spearman') {
                res$conf.int <- spearman_conf_int(res$estimate, res$n)
            }
        } else {
            res <- par_cor(m, s, 'bmi', data=dat, subset=subset, method=method)
        }
        cors[[i]][[m]][[s]] <- res
    }
}

# Create a table from results.
cors_table <- list()
for (group in groups) {
    ii <- grep(group, names(cors), value=TRUE)

    #prefix <- sub(paste0(group, '_'), '', ii)
    prefix <- sub('.+_', '', ii)
    headers <- c('cor', 'p', 'ci_lo', 'ci_hi')
    headers <- lapply(prefix, function(x) paste(x, headers, sep='_'))
    headers <- c('x', 'y', 'n', unlist(headers))

    tab <- data.frame()
    for (m in myokines) for (s in somato) {
        row <- list(m, s, cors[[ii[1]]][[m]][[s]]$n)
        for (i in ii) {
            res <- cors[[i]][[m]][[s]]
            row <- c(row, res$estimate, res$p.value, res$conf.int)
        }
        tab <- rbind(tab, unname(row))
    }
    colnames(tab) <- headers
    rownames(tab) <- NULL
    cors_table[[group]] <- tab
}
```

```{r print.table, rows.print=15}
cors_table$cases
cors_table$control
```

```{r save.table, eval=FALSE, include=FALSE}
dir.create('results', showWarnings=FALSE)
write.table(cors_table$cases, file='results/correlations-cases.tsv', quote=FALSE, sep='\t', row.names=FALSE)
write.table(cors_table$control, file='results/correlations-control.tsv', quote=FALSE, sep='\t', row.names=FALSE)
```

## Visualization {.tabset}

```{r plot}
pal <- brewer.pal(11, 'RdBu')[2:10]
color <- colorRampPalette(pal)(201)

cor_subplot <- function(x, y, correlations, ...) {
    current <- par('mfg')
    i <- current[1]
    j <- current[2]
    m <- current[3]
    n <- current[4]
    row <- n*(i - 1) + j

    # Get values from the 'correlations' table passed as argument.
    corr <- correlations[[i]][[j]]

    # Background color.
    idx <- round(100*(corr$estimate + 1) + 1)
    op <- par(usr=c(0, 1, 0, 1))
    symbols(.5, .5, rectangles=matrix(c(1, 1), nrow=1), bg=color[idx], fg=NULL, inches=FALSE, add=TRUE)
    par(op)

    # Data points and regression line.  Note that x and y are swapped.
    points(corr$y, corr$x, pch=20, cex=2.2, col='#00000060', ...)
    abline(lm(corr$x ~ corr$y), lwd=1.5)

    par(usr=c(0, 1, 0, 1))

    # Add legend.
    symbols(.5, .08, rectangles=matrix(c(1, .16), nrow=1), bg='white', fg=NULL, add=TRUE)
    #subtext <- sprintf("[%.2f, %.2f]", corr$conf.int[1], corr$conf.int[2])
    subtext <- sprintf("%.3f", corr$p.value)
    plot_text <- function(...) mtext(subtext, side=1, line=-1.1, cex=0.8, ...)
    if (corr$p.value < .05)
        plot_text(font=2, col='darkred')
    else
        plot_text(font=1, col='gray20')

    # Panel legend.
    panel_leg <- function(...) {
        mtext(..., line=2.5, las=2, cex=1)
    }
    # Topmost row.
    if (i == 1)
        panel_leg(somato[j], side=3)
    # Leftmost column.
    if (j == 1)
        panel_leg(myokines[i], side=2)
}

cor_plot <- function(data, main, ...) {
    m <- length(myokines)
    n <- length(somato)
    suppressWarnings(pairs(data, horInd=1:m, verInd=(m+1):(m+n), upper.panel=cor_subplot,
                           gap=0, oma=c(3, 15, 22, 3), ...))
    title(main, line=2.2, cex.main=1.5, adj=0.58)
}
```

### Cases - Unadjusted
```{r plot1}
main <- sprintf("Cases Samples - Spearmans's Correlation - Unadjusted")
cor_plot(dat[cases, -(1:2)], main, correlations=cors$cases_spearman_unadjusted)
```

### Control - Unadjusted
```{r plot2}
main <- sprintf("Control Samples - Spearmans's Correlation - Unadjusted")
cor_plot(dat[!cases, -(1:2)], main, correlations=cors$control_spearman_unadjusted)
```

### Cases - Adjusted
```{r plot3}
main <- sprintf("Cases Samples - Spearmans's Correlation - Adjusted by BMI")
cor_plot(dat[cases, -(1:2)], main, correlations=cors$cases_spearman_adjusted)
```

### Control - Adjusted
```{r plot4}
main <- sprintf("Control Samples - Spearmans's Correlation - Adjusted by BMI")
cor_plot(dat[!cases, -(1:2)], main, correlations=cors$control_spearman_adjusted)
```

##
```{r scale, fig.dim=c(10, 2)}
axis.args <- list(at=seq(-1, 1, by=0.25))
legend.scale(c(-1, 1), color, axis.args=axis.args)
```

```{r save.fig, eval=FALSE, include=FALSE}
for (group in groups) for (method in 'spearman') for (type in types) {
    result <- paste(group, method, type, sep='_')
    subset <- if (group == 'cases') cases else !cases
    main_type <- if (type == 'adjusted') paste(type, 'by BMI') else type
    main <- tools::toTitleCase(sprintf("%s's Correlations  –  %s  (%s)", method, main_type, group))
    png(sprintf('results/corr-%s-%s-%s.png', method, type, group), width=1000, height=1200, type='cairo')
    cor_plot(dat[subset, -(1:2)], main, correlations=cors[[result]])
    dev.off()
}

png('results/corr-color-scale-vertical.png', width=100, height=1020, type='cairo')
par(mar=c(2, 0, 2, 5))
legend.scale(c(-1, 1), color, axis.args=axis.args, horizontal=FALSE)
dev.off()
png('results/corr-color-scale-horizontal.png', width=1000, height=100, type='cairo')
par(mar=c(4, 2, 1, 2))
legend.scale(c(-1, 1), color, axis.args=axis.args)
dev.off()
```

<script>
    uncollapse = function() {
        $('.r-code-collapse')
            .filter((i, elem) => !/\n/.test(elem.textContent))
            .prev().find('button').click();
    }
    $(document).ready(() => setTimeout(uncollapse, 100))
</script>
